Historic collections as a tool for assessing the global pollination crisis

### 2018-06-13 revision/ms.Rmd
Revision (95%)

There is increasing concern about the decline of pollinators worldwide. However, despite reports that pollinator declines are widespread, data are scarce and often geographically and taxonomically biased. These biases limit robust inference about any potential pollinator crisis. 
Non-structured and opportunistic historical specimen collection data provide the only source of historical information which can serve as a baseline for identifying pollinator declines. Specimens historically collected and preserved in museums not only provide information on where and when species were collected, but also contain other ecological information such as species interactions and morphological traits. Here, we provide a synthesis of how researchers have used historical data to identify long-term changes in biodiversity, species abundances, morphology and pollination services. Despite recent advances, we show that information on the status and trends of most pollinators is absent, but we highlight opportunities and limitations to progress the assessment of pollinator declines globally. Finally, we demonstrate different approaches to analysing museum collection data using two contrasting case studies from distinct geographical regions (New Zealand and Spain) for which long-term pollinator declines have never been assessed. There is immense potential for museum specimens to play a central role in assessing the extent of the global pollination crisis.

@@

Animal pollinators are a critical component of both natural and agricultural ecosystems worldwide, given their role in plant reproduction [@Ollerton2011] and food security [@Klein2007]. As with many other taxa, pollinators are vulnerable to a range of anthropogenic disturbances, which can cause local and regional population declines or even extinctions. The vulnerability of pollinators was identified several decades ago, and was popularized in 1996 by the influential book “The forgotten pollinators” [@Buchmann1996]. However, early accounts of pollinator declines were somewhat anecdotal, given the lack of pollinator population data at that time. These initial claims triggered the first efforts to assess this potential issue and included the formation of a US National Academy of Science (NAS) panel in 2006, which was commissioned to assess the extent of pollinator declines. The NAS report concluded that “For most pollinator species […] the paucity of long-term population data and the incomplete knowledge of even basic taxonomy and ecology make definitive assessment of status exceedingly difficult” [@NationalResearchCouncil2007]. Since then, studies on pollinator responses to various global change drivers have multiplied rapidly. Researchers have now developed strong consensus that disturbances such as habitat destruction, land-use intensification, chemical exposure, exotic species and climate change are causing pollinator declines, and often act synergistically [@Goulson2015; @Potts2010]. Yet, the current status and population trends of most pollinator species worldwide remain unknown. For example, a recent IUCN report concluded that even for Europe’s comparatively well-studied bee fauna, greater than 55% of bee species fell into the “Data Deficient” category [@Nieto2014]. For countries outside of Europe and the US, data on pollinator populations is almost non-existent.

One of the main barriers to identifying long-term pollinator population trends is that pollinators are incredibly taxonomically diverse and include bees, flies, butterflies, beetles, birds, bats and lizards [@Ollerton2017]. Additionally, many pollinators are highly mobile, short-lived and small, which makes monitoring their populations difficult. Bees are generally regarded as the most important pollinator group due to their abundance, pollination efficiency and widespread distribution [@Winfree2011]. However, bees are diverse, with more than 20,000 species currently described worldwide, and often require expert taxonomists for identification. Furthermore, the uneven distribution of researchers has resulted in geographical biases in bee decline research [@Martin2012], as well as taxonomic biases toward species that are easier to identify, such as bumblebees [@Archer2014; @DePalma2016].

@@

At a global scale, current evidence of pollinator declines is highly limited with most data restricted to the US and Europe. It is unsurprising that studies on pollinator declines are biased towards developed western countries, which have also been subject to extensive anthropogenic disturbance. For example, in the UK and the Netherlands, a citizen science based study using both observations and museum collection data detected strong richness declines for bees, hoverflies and flowering plants [@Biesmeijer2006]. In the Netherlands, museum data have also revealed simultaneous plant and pollinator declines [@Scheper2014]. Specifically, bee species with the strongest host plant preferences (i.e., specialists) displayed the strongest declines and thus, were most threatened with extinction. However, it is important to note that even for these two countries, local estimates of pollinator richness are biased toward large cities and regions dominated by agriculture, and thus lack data for well-preserved natural areas. Further exploration of this dataset revealed that for declining pollinator taxa, the trend has attenuated in recent decades [@Carvalheiro2013].

Although studies of local pollinator communities often detect richness declines, regional richness may remain relatively stable. For example, regional estimates for bee species richness changes in the eastern US show moderate declines [@Bartomeus2013a] and very few regional extinctions [@Colla2012]. This is a pattern also detected in the UK, where relatively few regional bee extinctions has been reported [@Ollerton2014]. These regional findings are in stark contrast with the widespread local extinctions reported in local studies. For example, Burkle et al. [@Burkle2013] compared historical observations of bee species’ occurrences in a large forested ecosystem with remaining forest remnants and reports several local extinctions. However, it is important to note that there is strong concordance between local extinctions and regional declines [@Bartomeus2013c], suggesting that local extinctions are indicators of regional population declines.

Reported declines for bumblebees are the most severe of all pollinator taxa. For example, declines of up to 18% in local bumblebee richness have been reported for Belgium and the Netherlands [@Carvalheiro2013]. In other parts of Europe, local richness declines range from 5% in Great Britain [@Carvalheiro2013] to 42% in Denmark [@Dupont2011]. In the USA, reported bumble decline are also severe with estimates ranging between 25% [@Grixti2009] and 30% [@Bartomeus2013a]. However, studies on species richness changes for other pollinator taxa are both scarce and geographically restricted. For butterflies, the only evidence of richness declines comes from Europe. Butterfly species richness has declined substantially in the Netherlands and Belgium since the 1950’s, although declines in Great Britain have been less severe [@Carvalheiro2013]. In Belgium, another study [@Maes2001] found that richness declines have been severe (approximately 30%), although this study assessed richness changes over a longer time period (early 1900’s to 2000) compared to [@Carvalheiro2013] (1950-69 vs. 1970-80 and 1970-89 vs. 1990-2009). In parts of Germany, up to 70% declines in local butterfly richness have been reported [@Wenzel2006]. Compared with other insect pollinator taxa, there are very few studies on hoverfly species richness changes, which are all restricted to Europe. In Belgium, Great Britain and the Netherlands, hoverfly richness changes have been modest [@Carvalheiro2013]. In the Netherlands, moderate increases in hoverfly species richness have been shown, whereas in Great Britain no significant directional changes were detected [@Biesmeijer2006]. Furthermore, directionality (richness increase or decrease) varies depending on the time period assessed. For example, hoverfly richness decreased in Belgium by approximately 6% from 1950-69 to 1970-80, but increased by approximately 10% between 1970-89 and 1990-2009 [@Carvalheiro2013].

For ilustrative purposes, we mapped the findings of this studies in Figure 1 to show the strong contrast between bee species richness worldwide, whith bee diversity hotspots in mediterranean countries, against the paucity of countries for which we have any local or regional data on bee, syrphid or butterfly declines (raw data: Sup Mat 1). Despite outside of Europe and the US and for non-insect taxa, there are very few or no studies on pollinator declines using historical records, there are species-specific examples of historical losses from different parts of the world (e.g., *Bombus dalbhomi*; @Morales2013).

@@

Where digitization has been completed, the data provide a rich source of information, allowing assessment of the current status and long-term trends of pollinator populations [e.g., @Bartomeus2013a; @Scheper2014; @Young2017]. This is despite the fact that museum collections often have a number of biases, including unknown sampling effort, personal interests of collectors and the curatorial techniques used. For example, collectors tend to target rare or unusual over common taxa, discard damaged individuals or only accession a certain number of individuals. In addition, collections are often made opportunistically, leading to a spatial biases where difficult to access areas are under-sampled or conversely, where samples are biased towards easily accessed locations (e.g., towns/cities and/or roadsides). Further, museum collection data can only be used to determine where species are present and not where they are absent. However, given adequate sample sizes and appropriate statistical techniques, most biases can be accounted for [e.g., @Bartomeus2013a; @Keil2011; @Pearce2006].

@@

GBIF (https://www.gbif.org/) is a central repository for global species occurrence data. Much of these data come from museums, private collections and government research institutes, but several other sources are also integrated. In combination with the popular statistical language R [@RCoreTeam2017], GBIF can be directly queried into your computer [@Chamberlain2014] and data availability can be checked for the region of interest. Focusing on bee taxa, we show here the number of modern and historic bee records currently available for different countries (Figure 2a, Sup mat 2). Thirty-seven countries count with more than 1800 records in each time period, making these data potentially analyzable without further data collection effort (see Figure 2b and c for an initial exploration). However, a proper analysis of this dataset would need a careful look at the data as we detail later on with two examples (Spain and New Zealand). Most countries fall short in one or both axes. For example, a variety of countries located in different continents such as Switzerland, Sri Lanka, Nicaragua or Zimbawe have a decent number of recent records, but lack historical collections and researchers should prioritize the digitalization of old material before embarking on data analyses. It is also important to note that historical records are not always vouchered in local museums (i.e., many European and US museums contain large collections of pollinators from other countries). In contrast, more than 192 countries have less than 1000 records for both time periods, making them poor candidates for analysing long-term pollinator population trends. Aside from bees, similar exploratory analyses can easily be conducted for other taxa.

@@

**Figure 2:** Exploration of available data for bee records showing (a) The number of bee occurrences before 1980 and after 1980 in GBIF for each country. The upper right quadrat (records in black) contains well covered countries with New Zealand (NZ) and Spain (ES) marked in red. For those countries, we show a temptaive comparision of the rarefied number of species in both time periods and show that for most countries the number of species recorded is slightly lower (red lines) in recent time periods (b). A more careful analysis of this data would help complete the map of global declines (c). In the map we plot % change is species recorded in gbif for the available countries to show the geographic coverage. Note that this data are likely to contain strong undetected biases, as we explore below.

As stated above, once historical collection datasets are made available, researchers must identify any potential biases. We explore this process with two contrasting dataset examples (Spain and New Zealand). In the Spanish dataset, most of the data comes from a few specific locations and was collected by a few specific teams. Hence the geographical coverage is not representative. Even worst, historical and modern collections do not overlap spatially, making any inference imposible to interpret. In this case, contacted the original collectors of the historical data to define their sampling protocols. We then resurveyed the same sites (30 years after the original surveys) using the same sampling protocols. In contrast, the New Zealand dataset includes a wide suite of collectors and collection locations but shows no obvious biases in geographical and taxonomic coverage through time. We complemented gbif data with further museum collections for bees and flies and analyze the regional richness changes through time. For these two case studies, we provide annotated R scripts as examples of analysis for different dataset types (Sup Mat 3). These different analytical approaches allow us to reveal long-term trends in pollinator populations for regions with contrasting sampling histories. We hope this resource will encourage researchers to analyse data for regions where current information on pollinator declines is lacking.

@@

Spain provides an interesting study system because its natural habitats have been transformed extensively by humans over a long time period, but land-use is not as intensive compared with many other European countries. In addition, Spain is a bee diversity hotspot (Figure 1a) and maintains a relatively heterogeneous landscape. Spain has already digitalized a large amount of pollinator occurrence data for both historic and recent periods (Figure 2a). However, visual inspection of the data revealed clustering around a few localities. Further, historic records did not spatially match recent records, making comparisons difficult. For this dataset, most of the historic records were located around Valladolid and were collected by Enrique Asensio and collaborators. There has been no recent sampling of bees in this area. However, we found that Enrique systematically sampled six independent locations and that additional historical data were available at the “Museo de Historia Nacional” and other minor collections. Digitization of these records, along with a re-survey of the original sampling locations provided an excellent dataset for a before and after comparison of bee communities.

In brief, after cleaning taxonomic names for erratas and synonims using the taxize package [@Chamberlain2018], we checked for sampling completeness for both time periods and compared rarefied species richness for each site before and after 1980 with a paired t-test (rarefication at 1000 specimens). We found that there were a reduced number of species at sites after 1980 (mean difference 20.27 species; 95 percent confidence interval: -1.03, 41.58; t = 2.44, df = 5, P = 0.06). However, this trend was highly dependent on site identity, as two out of three sites showed no richness declines. Interestingly, these two localities were the two that has experienced less land use changes (both are natural areas embeded into agro-ecosystemss). In contrast the other 4 localities suffered large urban or agricutural intensification. In addition, species not further recorded in the re-survey are not a random selection of species, but are clustered in a few genera. For example, Andrenids and their parasites (e.g. Nomada) showed the strongest declines whereas Halictids tend to be more stable (Sup mat 4). This pattern of winners and loosers of land use intensification is in accordance with findings elsewhere [e.g. @Bartomeus2013a], indicating that some clades are more sensitive to disturbance than others.

@@

In global terms, human colonisation of New Zealand was relatively recent (c. 740 y) [@Wilmshurst2008]. Before human arrival, New Zealand was predominately forested, but has since been dramatically altered by people. Early Māori settlers cleared forests by burning and more recently, European colonists cleared large tracts of remaining forests and drained low-lying wetlands for agriculture, mostly before 1900 [@Perry2014]. Therefore, human activity likely affected pollinator communities in New Zealand long before surveys and specimen collections began. Nevertheless, we can use museum records to identify trends in pollinator communities during New Zealand’s more recent history.

@@

We followed protocols outlined in [@Bartomeus2013a] to analyse the New Zealand data at the regional level. First, we filtered our original datasets so that data used for analyses only included independent collection events. To do this, we removed specimens collected at the same location, on the same date, and by the same collector. We found our data had reasonable coverage across time periods, although there was a peak in collection occurrences from 1960-1980. Further exploration of the New Zealand native bee data raised doubts on collection completeness in records prior to 1970, so we removed these records from further analyses. We accounted for differences in collection effort through binning collection records by time so that each bin had a similar number of records but a different number of years. We then estimated richness for each time period bin by rarefying all bins to an equal number of specimens and calculated the mean species richness ±SE for each bin. Finally, we estimated the significance of change in richness using a permutation test that randomly reordered time periods and calculated the correlation between time period and species richness. Thus, reported P-values were the proportion of permutations that had higher or lower correlations compared to the correlation between richness and the actual chronological time period sequence.

Second, to determine if the probability of finding a species in the collection changed over time, we used a general linear model with a binomial distribution and a logit link. For species that showed overdispersion, we used a quasi-binomial distribution. Further, we only included species in this analysis for which we had 30 or more records. To account for differences in sampling effort between years, we weighted each year by the total number of samples collected that year.

@@

However, at the species level, we found that 11 out of 27 bee species increased in relative occurrence over time (10 native and one exotic) and three bee species declined in relative occurrence (one native and two exotic) (Figure 3). Interestingly, the two exotic bee species that declined in relative occurrence were both in the genus Bombus, which were intentionally introduced into New Zealand for the pollination of crops. Native bees that increased in relative occurrence were mostly from the genus Leioproctus, which are medium sized, ground nesting solitary bees. Only one out of 14 fly species increased in relative occurrence, which was exotic, whereas four species decreased in occurrence (three native and one exotic). Native flies that decreased in relative occurrence were all Syrphidae in the genus Helophilus.

@@

A recent study found that more than 90% of the papers investigating pollinator responses to land-use change focused solely on richness and abundance descriptors [@Winfree2011]. But in addition to local (alpha) diversity and regional (gamma) diversity, researchers need to assess changes in turnover between sites (beta diversity). Environmental changes often result in a few “winner” species and many “losers” species [@Bartomeus2013a]. Identifying winners and losers is critical as the few winners are often exotic and represent a subset of traits that facilitate survival in highly modified environments [@Stavert2017]. These changes can have important effects for pollination of native plant species and crops [@Stavert2018].

In addition, museum specimen collections can provide much more information besides species occurrence records, given that such information is recorded when digitizing collections. This is particularly important for identifying mechanisms of decline and adaptation. For example, recording the date of collection is particularly important for tracking of phenological advances congruent with contemporary climate change [@Bartomeus2011]. In addition, pollinator specimen labels often include information about the host plant on which the specimen was collected. This information critical for understanding past and present species interactions [@Tylianakis2008]. Aside from this information, bee specimens often contain pollen loads trapped on hairs, from which past visitation events can be identified [@Kleijn2008]. Finally, museum specimens can be measured to track evolutionary changes by measuring the traits of specimen. This approach has been already used to investigate tongue length [@Miller-Struttmann2015] and body size [@Oliveira2016] in response to climate and land-use change. Finally, plant herbariums can also contain indirect evidence of pollinator and pollination declines [60]. A basic information for basic for linking pollinator declines with its consequences for ecosystem functioning.

@@

Unleashing the power of museum collection data to answer pressing ecological and evolutionary questions is at our hands, but requires the coordinated effort of many actors. Using two case studies, we show that strong collaboration between museum curators and ecologists is key to understanding data and treating it appropriately. To progress our understanding of the global pollination crisis, researchers and curators must aim to digitize museum collection data and make it readily available in a format that is widely accessible. Centralization of regional and national museum collection data in existing global platforms, such as GBIF, would facilitate free and widespread access. However, datasets could also be stored in alternative webpages or database repositories (e.g., university and museum webpages or Dryad) providing they are thoroughly documented and easily retrieved and combined with other datasets using open science tools [e.g., @Dallas2016].

@@

To identify global trends in pollinator declines we require robust data, collected from diverse geographic regions. It is also crucial that these data are analysed appropriately. This requires researches to identify biases and to any fill taxonomic and geographic gaps where possible. We need to place increased emphasis on quantifying pollinator declines in regions outside of the US and Europe, and for pollinator groups other than bees. For the US and Europe, there have been few regional bee extinctions [@Ollerton2014; @Bartomeus2013a] but in disturbed ecosystems, declines are widespread [@DePalma2017; @Biesmeijer2006]. For most other pollinator taxa and regions throughout the world we know almost nothing. Moving forward, the first step for many taxa will be to identify and describe species. Only then can we begin to document pollinator declines.

>>>>

There is increasing concern about the decline of pollinators worldwide. However, despite reports that pollinator declines are widespread, data are scarce and often geographically and taxonomically biased. These biases limit robust inference about any potential pollinator crisis. 
Non-structured and opportunistic historical specimen collection data provide the only source of historical information which can serve as a baseline for identifying pollinator declines. Specimens historically collected and preserved in museums not only provide information on where and when species were collected, but also contain other ecological information such as species interactions and morphological traits. Here, we provide a synthesis of how researchers have used historical data to identify long-term changes in biodiversity, species abundances, morphology and pollination services. Despite recent advances, we show that information on the status and trends of most pollinators is absent. We highlight opportunities and limitations to progress the assessment of pollinator declines globally. Finally, we demonstrate different approaches to analysing museum collection data using two contrasting case studies from distinct geographical regions (New Zealand and Spain) for which long-term pollinator declines have never been assessed. There is immense potential for museum specimens to play a central role in assessing the extent of the global pollination crisis.

@@

Animal pollinators are a critical component of both natural and agricultural ecosystems worldwide, given their role in plant reproduction [@Ollerton2011] and food security [@Klein2007]. As with many other taxa, pollinators are vulnerable to a range of anthropogenic disturbances, which can cause local and regional population declines or even extinctions. The vulnerability of pollinators was identified several decades ago, and was popularized in 1996 by the influential book “The forgotten pollinators” [@Buchmann1996]. However, early accounts of pollinator declines were somewhat anecdotal, given the lack of pollinator population data at that time. These initial claims triggered the first efforts to assess this potential issue and included the formation of a US National Academy of Science (NAS) panel in 2006, which was commissioned to assess the extent of pollinator declines. The NAS report concluded that “For most pollinator species […] the paucity of long-term population data and the incomplete knowledge of even basic taxonomy and ecology make definitive assessment of status exceedingly difficult” [@NationalResearchCouncil2007]. Since then, studies on pollinator responses to various global change drivers have multiplied rapidly. Researchers have now developed strong consensus that disturbances such as habitat destruction, land-use intensification, chemical exposure, exotic species and climate change are causing pollinator declines, and often act synergistically [@Goulson2015; @Potts2010]. Yet, the current status and population trends of most pollinator species worldwide remain unknown. For example, a recent IUCN report concluded that even for Europe’s comparatively well-studied bee fauna, more than 55% of bee species fell into the “Data Deficient” category [@Nieto2014]. For countries outside of Europe and the US, data on pollinator populations is almost non-existent.

One of the main barriers to identifying long-term pollinator population trends is that pollinators are incredibly taxonomically diverse, including bees, flies, butterflies, beetles, birds, bats and lizards [@Ollerton2017]. Additionally, many pollinators are highly mobile, short-lived and small, which makes monitoring their populations difficult. Bees are generally regarded as the most important pollinator group due to their abundance, pollination efficiency and widespread distribution [@Winfree2011]. However, bees are diverse, with more than 20,000 species currently described worldwide, and often require expert taxonomists for identification. Furthermore, the worldwide uneven distribution of research studies has resulted in geographical biases in bee decline research [@Martin2012], as well as taxonomic biases toward species that are easier to identify, such as bumblebees [@Archer2014; @DePalma2016].

@@

At a global scale, current evidence of pollinator declines is highly limited with most data restricted to the US and Europe. It is unsurprising that studies on pollinator declines are biased towards developed western countries, which have also been subject to extensive anthropogenic disturbance. For example, in the UK and the Netherlands, a citizen science based study using both observations and museum collection data detected strong richness declines for bees, hoverflies and flowering plants [@Biesmeijer2006]. In the Netherlands, museum data have also revealed coupled plant and pollinator declines [@Scheper2014]. Specifically, bee species with the strongest host plant preferences (i.e., specialists) displayed the strongest declines and thus, were most threatened with extinction. However, it is important to note that even for these two countries, local estimates of pollinator richness are biased toward large cities and regions dominated by agriculture, and thus lack data for less human-developed areas [jamie check wording of the "human-developed"](). Further exploration of this dataset revealed that for declining pollinator taxa, the species richness declining trend has attenuated in recent decades [@Carvalheiro2013].

Although studies of local native pollinator communities often detect richness declines, regional native richness may remain relatively stable. For example, regional estimates for bee species richness changes in the eastern US show moderate declines [@Bartomeus2013a] and very few regional extinctions [@Colla2012]. This is a pattern also detected in the UK, where relatively few regional bee extinctions has been reported [@Ollerton2014]. These regional findings are in stark contrast with the widespread local extinctions reported in local studies. For example, Burkle et al. [@Burkle2013] compared historical observations of bee species’ occurrences in a large forested ecosystem with remaining forest remnants and reports several local extinctions. However, it is important to note that there is strong concordance between local extinctions and regional declines [@Bartomeus2013c], suggesting that local extinctions are indicators of regional population declines.

Reported declines for bumblebees are the most severe of all pollinator taxa. For example, declines of up to 18% in local bumblebee richness have been reported for Belgium and the Netherlands [@Carvalheiro2013]. In other parts of Europe, local richness declines range from 5% in Great Britain [@Carvalheiro2013] to 42% in Denmark [@Dupont2011] and 70 % of bumblebee species are clasified as threatened or with declining population trends [@Nieto2014]. In the USA, reported bumble decline are also severe with estimates ranging between 25% [@Grixti2009] and 30% [@Bartomeus2013a]. However, studies on species richness changes for other pollinator taxa are both scarce and geographically restricted. For butterflies, the only evidence of richness declines comes from Europe. Butterfly species richness has declined substantially in the Netherlands and Belgium since the 1950’s, although declines in Great Britain have been less severe [@Carvalheiro2013]. In Belgium, another study [@Maes2001] found that richness declines have been severe (approximately 30%), although this study assessed richness changes over a longer time period (early 1900’s to 2000) compared to [@Carvalheiro2013] (1950-69 vs. 1970-80 and 1970-89 vs. 1990-2009). In parts of Germany, up to 70% declines in local butterfly richness have been reported [@Wenzel2006]. Compared with other insect pollinator taxa, there are very few studies on hoverfly species richness changes, which are all restricted to Europe. In Belgium, Great Britain and the Netherlands, hoverfly richness changes have been modest [@Carvalheiro2013]. In the Netherlands, moderate increases in hoverfly species richness have been shown, whereas in Great Britain no significant directional changes were detected [@Biesmeijer2006]. Furthermore, directionality (richness increase or decrease) varies depending on the time period assessed. For example, hoverfly richness decreased in Belgium by approximately 6% from 1950-69 to 1970-80, but increased by approximately 10% between 1970-89 and 1990-2009 [@Carvalheiro2013].

For ilustrative purposes, we mapped the findings of these studies in Figure 1 to show the strong contrast between bee species richness worldwide, whith bee diversity hotspots in mediterranean countries, against the paucity of countries for which we have any local or regional data on bee, syrphid or butterfly declines (raw data: Sup Mat 1). Except for outside of Europe and the US and for non-insect taxa, there are very few or no studies on pollinator declines using historical records. However, there are species-specific examples of historical losses from different parts of the world (e.g., *Bombus dalbhomi*; @Morales2013).

@@

Where digitization has been completed, the data provide a rich source of information, allowing assessment of the current status and long-term trends of pollinator populations [e.g., @Bartomeus2013a; @Scheper2014; @Young2017]. This is despite the fact that museum collections often have a number of biases, including unknown sampling effort, personal interests of collectors and the curatorial techniques used. For example, collectors tend to target rare or unusual over common taxa, discard damaged individuals or only accession a certain number of individuals. In addition, collections are often made opportunistically, which may lead to spatial bias where difficult to access areas are under-sampled or conversely, where samples are biased towards easily accessed locations (e.g., towns/cities and/or roadsides). Further, museum collection data can only be used to determine where species are present and not where they are absent. However, given adequate sample sizes and appropriate statistical techniques, most biases can be accounted for [e.g., @Bartomeus2013a; @Keil2011; @Pearce2006].

@@

GBIF (https://www.gbif.org/) is a central repository for global species occurrence data. Much of these data come from museums, private collections and government research institutes, but several other sources are also integrated. In combination with the popular statistical language R [@RCoreTeam2017], GBIF can be directly queried into your computer [@Chamberlain2014] and data availability can be checked for the region of interest. Focusing on bee taxa, we show here the number of modern and historic bee records currently available for different countries (Figure 2a, Sup mat 2). We selected thirty-seven countries that count with more than 1800 records in each time period, making these data potentially analyzable without further data collection effort (see Figure 2b and c for an initial exploration). We selected this arbitrary threshold as a compromise between sample size and number of countries that can be included in a general analysis. However, a proper analysis of this dataset and the minimum sample size needed per country would need a careful look at the data characteristics as we detail later on with two examples (Spain and New Zealand). Most countries fall short in one or both axes. For example, a variety of countries located in different continents such as Switzerland, Sri Lanka, Nicaragua or Zimbawe have a decent number of recent records, but lack historical collections and researchers should prioritize the digitalization of old material before embarking on data analyses. It is also important to note that historical records are not always vouchered in local museums (i.e., many European and US museums contain large collections of pollinators from other countries). In contrast, more than 192 countries have less than 1000 records for both time periods, making them poor candidates for analysing long-term pollinator population trends. We focus on bees, but similar exploratory analyses can easily be conducted for other taxa.

@@

**Figure 2:** Exploration of available data for bee records showing (a) The number of bee occurrences before 1980 and after 1980 in GBIF for each country. The upper right quadrat (records in black) contains well covered countries with New Zealand (NZ) and Spain (ES) marked in red. For those countries, we show a temptaive comparision of the rarefied number of species in both time periods and show that for most countries the number of species recorded is slightly lower (red lines) in recent time periods (b). A more careful analysis of these data would help complete the map of global declines (c). In the map we plot % change is species recorded in gbif for the available countries to show the geographic coverage. Note that this data are likely to contain strong undetected biases, as we explore below.

As stated above, once historical datasets are made available, researchers must identify any potential biases. We explore this process with two contrasting dataset examples (Spain and New Zealand). In the Spanish dataset, most of the data comes from a few specific locations and was collected by a few specific teams. Hence the geographical coverage is not representative. Even worse, historical and modern collections do not overlap spatially, making any inference imposible to interpret. In this case, contacted the original collectors of the historical data to define their sampling protocols. We then resurveyed the same sites (30 years after the original surveys) using the same sampling protocols. In contrast, the New Zealand dataset includes a wide suite of collectors and collection locations but shows no obvious biases in geographical and taxonomic coverage through time. We complemented gbif data with further museum collections for bees and flies and analyze the regional richness changes through time. For these two case studies, we provide annotated R scripts as examples of analysis for different dataset types (Sup Mat 3). These different analytical approaches allow us to reveal long-term trends in pollinator populations for regions with contrasting sampling histories. We hope this resource will encourage researchers to analyse data for regions where current information on pollinator declines is lacking.

@@

Spain provides an interesting study system because its natural habitats have been transformed extensively by humans over a long time period, but land-use is not as intensive compared with many other European countries. In addition, Spain is a bee diversity hotspot (Figure 1a) and maintains a relatively heterogeneous landscape. Spain has already digitalized a large amount of pollinator occurrence data for both historic and recent periods (Figure 2a). However, visual inspection of the data revealed clustering around a few localities. Further, historic records did not spatially match recent records, making comparisons difficult. For this dataset, most of the historic records were located around Valladolid and were collected by Enrique Asensio and collaborators. There has been no recent sampling of bees in this area. However, we found that Enrique systematically sampled six independent locations and that additional historical data were available at the “Museo de Historia Nacional” and other minor collections (E. Asensio personal communication). Digitization of these records, along with a re-survey of the original sampling locations provided an excellent dataset for a before and after comparison of bee communities.

In brief, after cleaning taxonomic names for erratas and synonims using the taxize package [@Chamberlain2018], we checked for sampling completeness for both time periods and compared rarefied species richness for each site before and after 1980 with a paired t-test (sample size pre-1980: X; sample size pre-1980: X; rarefication at 1000 specimens). We found that there were a reduced number of species at sites after 1980 (mean difference 20.27 species; 95 percent confidence interval: -1.03, 41.58; t = 2.44, df = 5, P = 0.06). However, this trend was highly dependent on site identity, as two out of three sites showed no richness declines. Interestingly, these two localities were the two that has experienced less dramatic land use changes (both are natural areas embeded into agro-ecosystemss). In contrast the other 4 localities experienced large urban or agricutural intensification. In addition, species not further recorded in the re-survey are not a random selection of species, but are clustered in a few genera. For example, Andrenids and their parasites (e.g. Nomada) showed the strongest declines whereas Halictids tend to be more stable (Sup mat 4). This pattern of winners and loosers of land use intensification is in accordance with findings elsewhere [e.g. @Bartomeus2013a], indicating that some clades are more sensitive to disturbance than others.

@@

In global terms, human colonisation of New Zealand was relatively recent (c. 740 y) [@Wilmshurst2008]. Before human arrival, New Zealand was predominately forested, but has since been dramatically altered by people. We can use museum records to identify trends in pollinator communities during New Zealand’s more recent history.

@@

We followed protocols outlined in [@Bartomeus2013a] to analyse the New Zealand data. First, we filtered our original datasets so that data used for analyses only included independent collection events. To do this, we removed specimens collected at the same location, on the same date, and by the same collector. We found our data had reasonable coverage along time periods and across space, although there was a peak in collection occurrences from 1960-1980. Further exploration of the New Zealand native bee data raised doubts on collection completeness in records prior to 1970. For example, old specimens lack particular genus, and geographically restricted. Hence, we removed these records from further analyses. We accounted for differences in collection effort through binning collection records by time so that each bin had a similar number of records but a different number of years. We then estimated species richness for each time period bin by rarefying all bins to an equal number of specimens and calculated the rarefied species richness ±SE for each bin. Finally, we estimated the significance of change in richness using a permutation test that randomly reordered time periods and calculated the correlation between time period and species richness. Thus, reported P-values were the proportion of permutations that had higher or lower correlations compared to the correlation between richness and the actual chronological time period sequence.

Second, to determine if the probability of finding a species in the collection changed over time, we used a general linear model with a binomial distribution and a logit link. For species that showed overdispersion, we used a quasi-binomial distribution. Further, we only included species in this analysis for which we had 30 or more records. Preliminary analysis using random subsampling of specimens from species with large sample size indicated that below 30 records the trends become too variable. To account for differences in sampling effort between years, we weighted each year by the total number of samples collected that year.

@@

However, at the species level, we found that 11 out of 27 bee species increased in relative occurrence over time (10 native and one exotic) and three bee species declined in relative occurrence (one native and two exotic) (Figure 3). Interestingly, the two exotic bee species that declined in relative occurrence were both in the genus Bombus, which were intentionally introduced into New Zealand for the pollination of crops. Native bees that increased in relative occurrence were mostly from the genus Leioproctus, which are medium sized, ground-nesting, solitary bees. Only one out of 14 fly species increased in relative occurrence, which was exotic, whereas four species decreased in occurrence (three native and one exotic). Native flies that decreased in relative occurrence were all Syrphidae in the genus Helophilus.

@@

A recent study found that more than 90% of the papers investigating pollinator responses to land-use change focused solely on richness and abundance descriptors [@Winfree2011]. But in addition to local (alpha) diversity and regional (gamma) diversity, researchers need to assess changes in turnover between sites (beta diversity). Environmental changes often result in a few “winner” species and many “loser” species [@Bartomeus2013a]. Identifying winners and losers is critical as the few winners are often exotic and represent a subset of traits that facilitate survival in highly modified environments [@Stavert2017]. These changes can have important effects for pollination of native plant species and crops [@Stavert2018].

In addition, digitalized museum specimen collections can provide much more information besides species occurrence records, given that such information is recorded when digitizing collections. This is particularly important for identifying mechanisms of decline and adaptation. For example, recording the date of collection is particularly important for tracking of phenological advances congruent with contemporary climate change [@Bartomeus2011]. In addition, pollinator specimen labels often include information about the host plant on which the specimen was collected. This information critical for understanding past and present species interactions [@Tylianakis2008]. Aside from this information, bee specimens often contain pollen loads trapped on hairs, from which past visitation events can be identified [@Kleijn2008]. Finally, museum specimens can be measured to track evolutionary changes by measuring the traits of specimen. This approach has been already used to investigate tongue length [@Miller-Struttmann2015] and body size [@Oliveira2016, @Renauld2016] [jamie, ref to add]() in response to climate and land-use change. Finally, plant specimens stored in plant herbaria can also contain indirect evidence of pollinator and pollination declines [60] which can link pollinator declines with its consequences for ecosystem functioning.

@@

Unleashing the power of museum collection data to answer pressing ecological and evolutionary questions is at our hands, but requires the coordinated effort of many actors. Using two case studies, we show that strong collaboration between museum curators and ecologists is key to understanding data and treating it appropriately. To advance our understanding of the global pollination crisis, researchers and curators must aim to digitize museum collection data and make it readily available in a format that is widely accessible. Centralization of regional and national museum collection data in existing global platforms, such as GBIF, would facilitate free and widespread access. However, datasets could also be stored in alternative webpages or database repositories (e.g., university and museum webpages or Dryad) providing they are thoroughly documented and easily retrieved and combined with other datasets using open science tools [e.g., @Dallas2016].

@@

To identify global trends in pollinator declines we require robust data, collected from diverse geographic regions. It is also crucial that these data are analysed appropriately. This requires researches to identify biases and to fill any taxonomic and geographic gaps where possible. We need to place increased emphasis on quantifying pollinator declines in regions outside of the US and Europe, and for pollinator groups other than bees. For the US and Europe, there have been few regional bee extinctions [@Ollerton2014; @Bartomeus2013a] but in disturbed ecosystems, declines are widespread [@DePalma2017; @Biesmeijer2006]. For most other pollinator taxa and regions throughout the world we know almost nothing. Moving forward, the first step for many taxa will be to identify and describe species [REF]. Only then can we begin to document pollinator declines.
